A thermodynamic view of dusty protoplanetary disks

### 2017-05-23 appendix.tex

\subsection{Approximate solutions in the dust-rich limit}\label{si_dust_rich}
Here we seek analytic solutions for the streaming instability by 
examining limiting cases and with further approximations. We shall fix
$k_z$ and $\fdust>1/2$ (the dust-rich case), and maximize growth rates
over $k_x$. We consider small stopping times, $\tau_\mathrm{s}\to0$,
and assume that the optimum $k_x\to\infty$ but that
$k_x\tau_\mathrm{s}$ remain bounded.

We begin with the incompressible dispersion
relation, Eq. \ref{streaming_incompressible}. We assume a Keplerian
disk and neglect the quartic  
term because it is higher order in $\tau_\mathrm{s}$ (see also
\S\ref{spurious_epi})  In dimensionless form: 
\begin{align}
  \nu^3 + \st\left[\ii\tepsilon + 2 K_x \fg \left(1-\fdust\right)\right]\nu^2
  - \left(\frac{K_z}{K_x}\right)^2\nu - 2\st K_x \fg
  \left(1-2\fdust\right)\left(\frac{K_z}{K_x}\right)^2 = 0,\label{dimless_cubic}
\end{align}
where $\nu = \sigma/\Omega$, $\st = \taus\Omega$, $\fg = 1-\fdust$ (the gas
fraction), and recall $K_{x,z}=k_{x,z}\eta r$. We have used $F_r =
2\eta \fg r \Omega^2$ and considered large $K_x$.

Full analytic expressions for cubic roots are unwieldy. To proceed
in an a tractable manner, let us \emph{assume} at this stage that the
quadratic term can be neglected compare to the last term. This amounts
to assuming a  sufficiently \emph{large} $K_z$ such that :
\begin{align}
\left|\frac{\ii\fdust}{2K_x\fg} + \left(1 -
      \fdust\right)\right| \left|K_x\nu\right|^2 \ll
  K_z^2\left(2\fdust - 1\right) \label{min_kz}
\end{align}
This will be checked later. (For fixed $\fdust$ and $K_x\to\infty$, the
imaginary term can be neglected.)

Eq. \ref{dimless_cubic} now becomes the depressed cubic 
\begin{align}
  \nu^3 - \calP \nu = \calQ, \label{depressed_cubic}
\end{align}
where $\calP$ and $\calQ$ can be read off Eq. \ref{dimless_cubic}. This can be
solved with Vieta's substitution 
\begin{align*}
\nu = \mu + \frac{\calP}{3\mu},
\end{align*}
then Eq. \ref{depressed_cubic} becomes a quadratic for $\mu^3$,
\begin{align}
\mu^6 - \calQ\mu^3 + \frac{\calP^3}{27} = 0. 
\end{align}
The solution is
\begin{align}
  \mu^3 &= \frac{\calQ}{2}\left(1\pm \sqrt{1 -
      \frac{4\calP^3}{27\calQ^2}}\right)\notag\\
  & = \st \fg (1- 2\fdust)\frac{K_z^2}{K_x}\left\{1 \pm \sqrt{1 -
      \frac{K_z^2}{27K_x^2\left[\st K_x \fg
          \left(1-2\fdust\right)\right]^2}}\right\}.  
\end{align}
At this point we assume $K_z$ is sufficiently \emph{small} such that
the term $\propto K_z^2$ inside the square root may be neglected. That
is, 
\begin{align}
  K_z^2 \ll 27 \left(K_x^2\st\right)^2 \left[\fg
          \left(1-2\fdust\right)\right]^2. \label{max_kz}
%27 K_x^2 \left[\st K_x \fg
%          \left(1-2\fdust\right)\right]^2.
\end{align}
This will be checked later. Remembering that we are considering the
dust-rich case so that $1-2\fdust<0$, the solution for $\mu$ is   
\begin{align}
\mu = \left[2\st\fg \left(2\fdust
    -1\right)\frac{K_z^2}{K_x}\right]^{1/3}e^{\ii\pi/3}. \label{mu_explicit} 
\end{align}
We have chosen the root for instability. We can now write down the
solution for the eigenfrequency $\nu$:
\begin{align}
\real{(\nu)} &= \frac{1}{2}\left(|\mu| + \frac{K_z^2}{3
    K_x^2|\mu|}\right)\notag\\
&= \frac{1}{2}\left(\calA \st^{1/3}K_x^{-1/3} + \calB
    \st^{-1/3}K_x^{-5/3}\right),\\
\imag{(\nu)} &= \frac{\sqrt{3}}{2}\left(|\mu| - \frac{K_z^2}{3
    K_x^2|\mu|}\right),\notag\\
&= \frac{\sqrt{3}}{2}\left(\calA \st^{1/3}K_x^{-1/3} - \calB
    \st^{-1/3}K_x^{-5/3}\right),
\end{align}
%Inserting the expression for $|\mu|$ into $\imag{\nu}$, the
%explicit expression for the growth rate may be written as 
%\begin{align}
%  s = \frac{\sqrt{3}}{2}\left(\calA \st^{1/3}K_x^{-1/3} - \calB
%    \st^{-1/3}K_x^{-5/3}\right)\OmK, 
%\end{align}
where
\begin{align}
  &\mathcal{A} = \left[2\fg \left(2\fdust
      -1\right)K_z^2\right]^{1/3} ,\\
  &\mathcal{B} = \frac{K_z^2}{3\calA}. 
\end{align}

Maximizing growth rates over $K_x$ by setting $\p \imag{(\nu)}/\p K_x
= 0$, we find the optium wavenumber is given by
\begin{align}
  K_x = \left(\frac{5\calB}{\calA}\right)^{3/4}\st^{-1/2} =
  \left(\frac{5}{3}\right)^{3/4}\sqrt{\frac{K_z}{2\fg\left(2\fdust-1\right)\st}}. 
\end{align}
The maximum growth rate is then 
\begin{align}
  \mathrm{max}\left(s\right) =
  \frac{2\sqrt{3}\calA^{5/4}}{5^{5/4}\calB^{1/4}}\sqrt{\st}\OmK =
  \frac{2\sqrt{2}\times3^{3/4}}{5^{5/4}}\sqrt{\fg\left(2\fdust-1\right)K_z\st}\OmK.
\end{align}
and the corresponding (dimensional) real frequency satisfies
$\real{(\sigma)} =  \sqrt{3}s/2$.

From the above result we see that as $\taus\to0$, the most unstable
radial wavelength diverges, $K_x\propto \taus^{-1/2} \to \infty$ with
a vanishing growth rate, $\mathrm{max}(s)\propto \sqrt{\taus}\to 0$.

\subsubsection{Finite phase lag as $\taus\to 0$}  
An interesting property of the special solutions described above is
that there is always a phase lag between the Lagrangian pressure and
density perturbations. Since $\real{(\nu)}>0$, the phase lag may be
defined as $\varphi = \arg{\left(\Delta P \Delta \rho^*\right)}$. For
SI we neglect the background density gradients, so $\Delta\rho = 
\delta\rho$. It may be shown that in the limit of large $K_x$ 
that $\Delta P \simeq \ii \delta v_x \p_rP/\sigma$. To obtain $\dd v_x$ we use Eq. \ref{streaming_vx} 
assuming low frequency modes, $|\sigma|\ll \kappa^2$, valid for
$\taus\to 0$ as shown above. Now, the expression for $\dd v_x$ involves
the Eulerian pressure perturbation, $\delta P$. For this we use
Eq. \ref{si_QW1} and assume incompressibility.

With these simplifications (in addition to those already made in the
previous section), the phase lag $\varphi$ is given via 
\begin{align}
  \tan{\varphi} = \frac{\imag{\left(\Delta
        P\Delta\rho^*\right)}}{\real{\left(\Delta
        P\Delta\rho^*\right)}}\simeq\frac{K_x\imag{(\nu)}}{2\fg^2\st
      K_x^2 + K_x \real{(\nu)}}. 
\end{align}
From here it is clear for the solutions derived above, where
$K_x\propto \st^{-1/2}$ and $\nu \propto \st^{1/2}$, that $\varphi$ is
a constant. Explicitly inserting the solutions give 
\begin{align}
\tan{\varphi} = \frac{2\times 3^{3/2}\left(2\fdust-1\right)}{16 -
  7\fdust}. 
\end{align}
For the case shown in Fig. \ref{si_compare_fig} with $\fdust=0.75$, we
have $\varphi\simeq 26\degr$, comparable to the $\sim 30\degr$
obtained from numerical solutions.

\subsubsection{Validty of the approximations}
The solution above require $K_z$ be sufficiently large for
Eq. \ref{min_kz} but also sufficiently small for
Eq. \ref{max_kz}. Inserting the above solutions, we find the
requirement from Eq. \ref{min_kz} becomes 
\begin{align}
  \fdust \gg \frac{12}{17}\simeq 0.7.
\end{align}
However, since $\fdust \leq 1$, this requirement can only be marginally
satisfied. We find that this mostly give errors in the real
frequency (see Fig. \ref{si_compare_fig}), while growth rates are
still captured. On the other hand, Eq. \ref{max_kz} becomes the trivial 
inequality \begin{align}
  1\ll \frac{27}{4}\left(\frac{5}{3}\right)^3\simeq 30,
\end{align}
so Eq. \ref{max_kz} is satisified, which justifies the solution for
$\mu$ in Eq. \ref{mu_explicit}.

>>>>

\subsection{Approximate solutions in the dust-rich limit}\label{si_dust_rich}
Here we seek analytic solutions for the streaming instability by 
examining limiting cases and with additional approximations. We will
fix $k_z, \fdust$ and maximize growth rates over $k_x$. It turns out
simple solutions exist in the dust-rich case with 
$\fdust$ near unity. We consider small stopping times, 
$\tau_\mathrm{s}\to0$, and assume that the corresponding optimum
$k_x\to\infty$.

We begin with the incompressible dispersion
relation, Eq. \ref{streaming_incompressible}. We assume a Keplerian
disk ($\Omega=\kappa=\OmK$), and low frequency  modes 
($|\sigma|\ll \OmK$), which leads to neglecting the quartic  
term. (This is also necessary to avoid spurious modes, see
\S\ref{spurious_epi}.)  In dimensionless form:   
\begin{align}
  \nu^3 + \st\left[\ii\tepsilon + 2 K_x \fg \left(1-\fdust\right)\right]\nu^2
  - \left(\frac{K_z}{K_x}\right)^2\nu - 2\st K_x \fg
  \left(1-2\fdust\right)\left(\frac{K_z}{K_x}\right)^2 = 0,\label{dimless_cubic}
\end{align}
where $\nu = \sigma/\OmK$, $\st = \taus\OmK$, $\fg = 1-\fdust$ (the gas
fraction), and recall $K_{x,z}=\eta r k_{x,z}$. We have used $F_r =
2\eta \fg r \Omega^2$ and considered large $K_x$.

Analytic expressions for cubic roots are unwieldy. To keep the problem
tractable, let us \emph{assume} at this stage that the quadratic term
can be neglected compare to the last term. That is, 
%This amounts 
%to assuming a  sufficiently \emph{large} $K_z$ such that :
\begin{align}
\left|\frac{\ii\fdust}{2K_x\fg} + \left(1 -
      \fdust\right)\right| \left|K_x\nu\right|^2 \ll
  K_z^2\left(2\fdust - 1\right). \label{min_kz}
\end{align}
(For fixed $\fdust$ but allowing $K_x\to\infty$, the imaginary term can be
neglected.) We show in \S\ref{si_dust_rich_approx} that this
simplification is valid for dust-rich disks.

Eq. \ref{dimless_cubic} now becomes the depressed cubic 
\begin{align}
  \nu^3 - \calP \nu = \calQ, \label{depressed_cubic}
\end{align}
where $\calP$ and $\calQ$ can be read off Eq. \ref{dimless_cubic}. This can be
solved with Vieta's substitution 
\begin{align}
\nu = \mu + \frac{\calP}{3\mu},\label{si_vieta}
\end{align}
then Eq. \ref{depressed_cubic} becomes a quadratic for $\mu^3$,
\begin{align}
\mu^6 - \calQ\mu^3 + \frac{\calP^3}{27} = 0. 
\end{align}
The solution is
\begin{align}
  \mu^3 = \frac{\calQ}{2}\left(1\pm \sqrt{1 -
      \frac{4\calP^3}{27\calQ^2}}\right) 
   = \st \fg (1- 2\fdust)\frac{K_z^2}{K_x}\left\{1 \pm \sqrt{1 -
      \frac{K_z^2}{27K_x^2\left[\st K_x \fg
          \left(1-2\fdust\right)\right]^2}}\right\}.  
\end{align}
At this point we assume the term $\propto K_z^2$ inside the square
root may be neglected. That is, 
\begin{align}
  K_z^2 \ll 27 \left(K_x^2\st\right)^2 \left[\fg
          \left(1-2\fdust\right)\right]^2. \label{max_kz}
%27 K_x^2 \left[\st K_x \fg
%          \left(1-2\fdust\right)\right]^2.
\end{align}
We show in \S\ref{si_dust_rich} that this is readily satisfied. Then
$\mu \simeq \mathcal{Q}^{1/3}$.

Now, remembering that we are considering the dust-rich limit with 
$1-2\fdust<0$, the explicit solution for $\mu$ is    
\begin{align}
\mu = \left[2\st\fg \left(2\fdust
    -1\right)\frac{K_z^2}{K_x}\right]^{1/3}e^{\ii\pi/3}. \label{mu_explicit} 
\end{align}
We have chosen the complex root for instability. Inserting this into
Eq. \ref{si_vieta} gives the eigenfrequency $\nu$:
\begin{align}
\real{(\nu)} %&= \frac{1}{2}\left(|\mu| + \frac{K_z^2}{3
%    K_x^2|\mu|}\right)\notag\\
&= \frac{1}{2}\left(\calA \st^{1/3}K_x^{-1/3} + \calB
    \st^{-1/3}K_x^{-5/3}\right),\\
\imag{(\nu)} %&= \frac{\sqrt{3}}{2}\left(|\mu| - \frac{K_z^2}{3
%    K_x^2|\mu|}\right),\notag\\
&= \frac{\sqrt{3}}{2}\left(\calA \st^{1/3}K_x^{-1/3} - \calB
    \st^{-1/3}K_x^{-5/3}\right),
\end{align}
%Inserting the expression for $|\mu|$ into $\imag{\nu}$, the
%explicit expression for the growth rate may be written as 
%\begin{align}
%  s = \frac{\sqrt{3}}{2}\left(\calA \st^{1/3}K_x^{-1/3} - \calB
%    \st^{-1/3}K_x^{-5/3}\right)\OmK, 
%\end{align}
where
\begin{align}
  &\mathcal{A} = \left[2\fg \left(2\fdust
      -1\right)K_z^2\right]^{1/3} ,\\
  &\mathcal{B} = \frac{K_z^2}{3\calA}. 
\end{align}

Maximizing growth rates over $K_x$ by setting $\p \imag{(\nu)}/\p K_x
= 0$, we find the optium wavenumber is given by
\begin{align}
  K_x = \left(\frac{5\calB}{\calA}\right)^{3/4}\st^{-1/2} =
  \left(\frac{5}{3}\right)^{3/4}\sqrt{\frac{K_z}{2\fg\left(2\fdust-1\right)\st}}. \label{si_dust_rich_optkx}
\end{align}
The maximum growth rate is then 
\begin{align}
  \mathrm{max}\left[\imag{(\nu)}\right] =
  \frac{2\sqrt{3}\calA^{5/4}}{5^{5/4}\calB^{1/4}}\sqrt{\st} =
  \frac{2\sqrt{2}\times3^{3/4}}{5^{5/4}}\sqrt{\fg\left(2\fdust-1\right)K_z\st}.\label{si_dust_rich_grow}
\end{align}
and the real frequency $\real{(\nu)} =
(\sqrt{3}/2)\imag{(\nu)}$. We see that as $\taus\to0$, the most unstable
radial wavenumber diverges, $K_x\propto \taus^{-1/2} \to \infty$ with 
a vanishing growth rate, $\mathrm{max}(s)\propto \sqrt{\taus}\to
0$. This corresponds to instability at arbitrarily small radial
lengthscales.

\subsubsection{Finite phase lag as $\taus\to 0$}  
An interesting property of the special solutions described above is
that there is always a phase lag between the Lagrangian pressure and
density perturbations. Since $\real{(\nu)}>0$, the phase lag may be
defined as $\varphi = \arg{\left(\Delta P \Delta \rho^*\right)}$. %For
%SI we neglect the background density gradients, so $\Delta\rho
%\simeq  \delta\rho$ but retain the background pressure gradient. 
(Note that we may set $\Delta\rho = 1$ without loss of generality.) 
It may be shown that in the limit of large $K_x$  that $\Delta P
\simeq \ii \delta v_x \p_rP/\sigma$. To obtain $\dd v_x$ we use
Eq. \ref{streaming_vx} in the low frequency limit. The expression for $\dd v_x$ involves
the Eulerian pressure perturbation, $\delta P$, for which we use 
Eq. \ref{si_QW1} and assume incompressibility ($\zeta\to0$).

With these additional simplifications the phase lag $\varphi$ is given
via  
\begin{align}
  \tan{\varphi} = \frac{\imag{\left(\Delta
        P\Delta\rho^*\right)}}{\real{\left(\Delta
        P\Delta\rho^*\right)}}\simeq\frac{K_x\imag{(\nu)}}{2\fg^2\st
      K_x^2 + K_x \real{(\nu)}}. 
\end{align}
From here it is clear for the above solutions, where
$K_x\propto \st^{-1/2}$ and $\nu \propto \st^{1/2}$, that $\varphi$ is
a constant. Explicitly inserting the solutions give 
\begin{align}
\tan{\varphi} = \frac{2\times 3^{3/2}\left(2\fdust-1\right)}{16 -
  7\fdust}. 
\end{align}
For the case shown in Fig. \ref{si_compare_fig} with $\fdust=0.75$, we
have $\varphi\simeq 26\degr$, comparable to the $\sim 30\degr$
obtained from numerical solutions.

\subsubsection{Validty of the approximations}\label{si_dust_rich_approx}
Inserting the solutions for the most unstable mode 
%(Eq. \ref{si_dust_rich_optkx}---\ref{si_dust_rich_grow} with ) 
we find the assumption in Eq. \ref{min_kz} becomes  
\begin{align}
  \fdust \gg \frac{12}{17}\simeq 0.7.
\end{align}
Since $\fdust \leq 1$, this requirement can only be marginally
satisfied. However, we find that this mostly give errors in the real
frequency (see Fig. \ref{si_compare_fig}), while growth rates are
still captured correctly. 
On the other hand, the assumption of Eq. \ref{max_kz} becomes the trivial
inequality \begin{align} 
  1\ll \frac{27}{4}\left(\frac{5}{3}\right)^3\simeq 30,
\end{align}
so Eq. \ref{max_kz} is satisified, which justifies the approximate solution for 
$\mu$ in Eq. \ref{mu_explicit}.
